### 2016-09-15 GPCE2016/sections/Introduction.tex
Introduction revised

\Objectoriented languages strive to offer great code reuse.
They couple flexibility and rigour, expressive power and
modular reasoning.  Two main ideas emerged to this end:
\prototypebased (PB)~\cite{Ungar87self} and \classbased languages such as
Java, C\# or Scala.  In \prototypebased
languages objects inherit from other objects. Thus objects own
both behaviour and state (and objects are all you have).
In \classbased languages an object is an instance of a specific class,
and classes inherit from other classes.  Objects own state,
while classes contain behaviour and the structure of the state.

This paper presents a third alternative: the concept of
\emph{\interfacebased} \objectoriented programming languages (IB), where objects
implement interfaces directly. In IB interfaces own the implementation
for the behaviour, which is structurally defined in their
interface. Programmers do not define objects directly but delegate
the task to \emph{object interfaces}, whose role is similar to non-abstract
classes in \classbased \objectoriented programming languages
(CB). Objects
are instantiated by static factory methods in object interfaces.

A key challenge in IB lies in how to model state, which is
fundamental to have stateful objects. All abstract operations in an
object interface are interpreted as \emph{abstract state
  operations}. The abstract state operations include various common
utility methods (such as getters and setters, or clone-like
methods). Objects are only responsible to define the ultimate
behaviour of a method. % and, for example, if such a method is just a setter.
Anything related to state is completely contained in the
instances and does not leak into the inheritance logic.  In CB, the structure of the state is fixed and can only grow
by inheritance.  In contrast, in IB the state is never
fixed, and methods such as abstract setters and getters
can always receive an explicit implementation down in the inheritance
chain, improving \textbf{modularity and flexibility}.  That is, the
concept of abstract state is more fluid.

@@

One advantage of abstract state operations and
type-refinement is that it allows a new approach to
\emph{type-safe covariant mutable state}. That is, in IB,
it is possible to type-refine \emph{mutable} ``fields'' in subtypes. This is
typically forbidden in CB: it is widely known that \emph{naive} type-refinement of
mutable fields is not type-safe. Although covariant refinement of mutable
fields is supported by some type systems~\cite{bruce98astatically,bruce1994paradigmatic,ernst06virtual,Saito2013933}, this requires
significant complexity and restrictions to ensure that all uses of
covariant state are indeed type-safe.

@@

Another advantage of IB is the support for
multiple inheritance. The literature shows how
easy it is to combine multiple sources of pure behaviour
using traits~\cite{scharli03traits}. In Java
multiple \emph{interface} inheritance has been supported since
inception, and Java 8 default methods~\cite{goetz12fdefenders} bring some of the
advantages of traits to Java. The literature~\cite{Sak89dis,bracha90mixin,malayeri2009cz}
is also rich on how hard it is to combine multiple sources
of behaviour \textbf{and} state with multiple \emph{implementation}
inheritance of classes. In IB there is only multiple
interface inheritance, yet programmers can still use state via
abstract state operations. IB enables powerful
idioms using multiple inheritance and state.

@@

IB could be explained by defining a novel language, with new syntax
and semantics. However, this would have a steep learning curve.  We
take a different approach instead. For the sake of providing a more
accessible explanation, we will embed our ideas directly into Java.
Our IB embedding relies on the
new features of Java 8: interface \emph{static methods} and
\emph{default methods}, which allow interfaces to have method
implementations. In the context of Java, what we propose is a programming
style, where we never use classes (more precisely, we never
  use the \Q@class@ keyword).  We call this restricted version of
Java \emph{Classless Java}. \textcolor{red}{Classless Java provides us
with a way to address the famous diamond problem~\cite{bracha90mixin,Sak89dis}
based on Java 8's interface model: interfaces can offer different implementations
using the default methods, whereas conflicts are supposed to be explicitly resolved,
otherwise the program will be rejected by the compiler. An interface that inherits
a few default implementations of a method from its supertypes can define a new implemention
to hide the conflicts among them, or use the \Q@super@ keyword to specify the one it wants. Examples
and more discussion will be presented afterwards.}

@@

Using Java annotation processors, we produce an implementation of
Classless Java, which allows us to stick to pure Java 8. By annotating
 the interfaces that represent object interfaces with \mixin, code for interface instantiation and
type refinement can be automatically generated. Such code should not be needed in the first place in
a real IB language, and the annotation processor allows us to
transparently hide it from Java programmers. The implementation works
by performing AST rewriting, allowing most existing Java tools (such as
IDEs) to work out-of-the-box with our implementation. Moreover, the
implementation blends Java's conventional CB style and IB smoothly.
As a result, we apply object interfaces to several interesting
Java programs and conduct various case studies. Finally, we also
discuss the behaviour of our \mixin annotation and
its properties.

@@

While the Java embedding has obvious advantages from the practical
point-of-view, it also imposes some limitations that a new IB language
would not have. Supporting proper encapsulation is
difficult in Java due to limitations of Java interfaces. In particular
in Java interfaces the visibility of all methods is public. Thus
modelling private state is difficult in current Java 8. However,
 using existing design patterns~\cite{BettiniDSS13} we can emulate hiding
methods from interfaces. Furthermore Java 9 will allow private methods
in interfaces~\cite{srikanth16private}.

>>>>

\Objectoriented languages strive to offer great code reuse.
They couple flexibility and rigour, expressive power and
modular reasoning.  Two main OO models emerged to this end:
\prototypebased (PB)~\cite{Ungar87self} and \classbased languages such as
Java, C\# or Scala.  In \prototypebased
languages objects inherit from other objects. Thus objects own
both behaviour and state (and objects are all you have).
In \classbased languages an object is an instance of a specific class,
and classes inherit from other classes.  Objects own state,
while classes contain behaviour and the structure of the state.

Regardeless of the OO model, inheritance is a key mechanism is OO 
languages. Inheritance provides a modularization mechanism, which 
is used to reuse implementations from inherited classes/objects. 
Unfortunatelly, as widely acknowledged in the literature~\cite{scharli03traits,Sak89dis,bracha90mixin,malayeri2009cz}
multiple inheritance, especially when combined with state, has several
tricky issues, including several variants of the famous 
\emph{diamond problem}~\cite{}. Many of the problems related to 
inheritance arize from the direct use of \emph{fields} to model state. 
Inheriting two fields of the same name raises 
the question of whether the two fields should be kept, or only one 
field should exists (and if so, which value should be picked). Initialization 
of the fields is also problematic, since initialization code may be inherited from 
multiple parents. Finally, an additional problem with mutable fields is that their 
type cannot be type-refined in extensions, which can cause modularity problems
~\cite{wadler98expression,eptrivially}.

To address those limitations, this paper presents a third alternative
OO model called \emph{\interfacebased} \objectoriented programming
languages (IB), where objects implement interfaces directly and fields
are not directly supported. In IB interfaces own the implementation for the
behaviour, which is structurally defined in their
interface. Programmers do not define objects directly but delegate the
task to \emph{object interfaces}, whose role is similar to
non-abstract classes in \classbased \objectoriented programming
languages (CB). Objects are instantiated by static factory methods in
object interfaces.

Due to the absence of fields, a key challenge in IB lies in how to model state, which is
fundamental to have stateful objects. All abstract operations in an
object interface are interpreted as \emph{abstract state
  operations}. The abstract state operations include various common
utility methods (such as getters and setters, or clone-like
methods). Objects are only responsible to define the ultimate
behaviour of a method. % and, for example, if such a method is just a setter.
Anything related to state is completely contained in the
instances and does not leak into the inheritance logic.  In CB, the structure of the state is fixed and can only grow
by inheritance.  In contrast, in IB the state is never
fixed, and methods such as abstract setters and getters
can always receive an explicit implementation down in the inheritance
chain, improving \textbf{modularity and flexibility}.  That is, the
concept of abstract state is more fluid.

@@

One advantage of abstract state operations and
type-refinement is that it allows a new approach to
\emph{type-safe covariant mutable state}. That is, in IB,
it is possible to type-refine \emph{mutable} state in subtypes. This is
typically forbidden in CB: it is widely known that \emph{naive} type-refinement of
mutable fields is not type-safe. Although covariant refinement of mutable
fields is supported by some type systems~\cite{bruce98astatically,bruce1994paradigmatic,ernst06virtual,Saito2013933}, this requires
significant complexity and restrictions to ensure that all uses of
covariant state are indeed type-safe.

@@

\begin{comment}
Another advantage of IB is the support for
multiple inheritance. The literature shows how
easy it is to combine multiple sources of pure behaviour
using traits~\cite{scharli03traits}. In Java
multiple \emph{interface} inheritance has been supported since
inception, and Java 8 default methods~\cite{goetz12fdefenders} bring some of the
advantages of traits to Java. The literature~\cite{Sak89dis,bracha90mixin,malayeri2009cz}
is also rich on how hard it is to combine multiple sources
of behaviour \textbf{and} state with multiple \emph{implementation}
inheritance of classes. In IB there is only multiple
interface inheritance, yet programmers can still use state via
abstract state operations. IB enables powerful
idioms using multiple inheritance and state.
\end{comment}

@@

IB could be explained by defining a novel language, with new syntax
and semantics. However, this would have a steep learning curve.  We
take a different approach instead. For the sake of providing a more
accessible explanation, we will embed our ideas directly into Java.
Our IB embedding relies on the
new features of Java 8: interface \emph{static methods} and
\emph{default methods}, which allow interfaces to have method
implementations. In the context of Java, what we propose is a programming
style, where we never use classes (more precisely, we never
  use the \Q@class@ keyword).  We call this restricted version of
Java \emph{Classless Java}. 
\begin{comment}
\textcolor{red}{Classless Java provides us
with a way to address the famous diamond problem~\cite{bracha90mixin,Sak89dis}
based on Java 8's interface model: interfaces can offer different implementations
using the default methods, whereas conflicts are supposed to be explicitly resolved,
otherwise the program will be rejected by the compiler. An interface that inherits
a few default implementations of a method from its supertypes can define a new implemention
to hide the conflicts among them, or use the \Q@super@ keyword to specify the one it wants. Examples
and more discussion will be presented afterwards.}
\end{comment}

@@

Using Java annotation processors, we produce an implementation of
Classless Java, which allows us to stick to pure Java 8. 
%By annotating
% the interfaces that represent object interfaces with \mixin, code for interface instantiation and
%type refinement can be automatically generated. Such code should not be needed in the first place in
%a real IB language, and the annotation processor allows us to
%transparently hide it from Java programmers. 
The implementation works
by performing AST rewriting, allowing most existing Java tools (such as
IDEs) to work out-of-the-box with our implementation. Moreover, the
implementation blends Java's conventional CB style and IB smoothly.
We apply object interfaces to several interesting
Java programs and conduct various case studies. Finally, we also
discuss the behaviour of Classless Java and
its properties.

@@

\begin{comment}
While the Java embedding has obvious advantages from the practical
point-of-view, it also imposes some limitations that a new IB language
would not have. Supporting proper encapsulation is
difficult in Java due to limitations of Java interfaces. In particular
in Java interfaces the visibility of all methods is public. Thus
modelling private state is difficult in current Java 8. However,
 using existing design patterns~\cite{BettiniDSS13} we can emulate hiding
methods from interfaces. Furthermore Java 9 will allow private methods
in interfaces~\cite{srikanth16private}.
\end{comment}
